A Simplified Plackett-Luce Likelihood Function for Rank Orders

### 2024-09-03 inst/hankin-AustrianJournalofStatistics.tex
fixes #240

Here a simplification of the Plackett--Luce likelihood functions for
order statistics is proposed.  We consider a field of $n+1$ entities
that are ordered in some way [the preferred example is competitors in
  a running race, ordered by time of crossing the finishing line].
One entity is of particular interest to us: the ``focal entity'', or
focal competitor.  We assign Plackett--Luce strength $a$ to the focal
competitor and $b$ to all others; we require $a+b=1$.  The associated
inference problem has a wide range of applications and I present some
analysis of datasets drawn from the fields of Olympic athletics,
education, and Formula~1 motor racing.

}
\Keywords{keywords, comma-separated, non-capitalized, \proglang{R}}
\Plainkeywords{keywords, comma-separated, non-capitalized,, R} %% without formatting
%% at least one keyword must be supplied

@@

As noted, likelihood function~\ref{plackettluce} becomes unmanageable
with $n$ even moderately large and some form of simplification is
needed.  To that end, consider a Plackett-Luce likelihood function for
a race between a competitor of Bradley-Terry strength $a$ and
$n\geqslant 1$ cloned competitors, each of strength $b=1-a$.  Under
these circumstances, an observation is indexed by~$r$, the number of
strength $b$ clones finishing ahead of~$a$, and because we have $n+1$
competitors $0\leqslant r\leqslant n$.  The initial field strength is
$a+nb=n-(n-1)a$.  The Plackett--Luce likelihood
function~\citep{luce1959,plackett1975} for this situation would be

\begin{eqnarray}\label{likeforrn1}
  \mathcal{L}_{r,n}(a) &=&
\frac{b}{a+ n   b}\cdot
\frac{b}{a+(n-1)b}\cdots\frac{a}{a+(n-r)b}\nonumber\\
&=& \frac{B-1}{(B+n-1)(B+n-2)\cdots(B+n-r-1)}\nonumber\\ 
&=& \frac{(B-1)(B+n-r-2)!}{(B+n-1)!}\label{likeforrn3}
\end{eqnarray}

\noindent where $B=1/b$.  Note carefully that these expressions are
likelihoods, not probabilities, and do not sum to 1.  Previous related
work \citep{hankin2024_hyper3} presented the case $n=2$ [in current
  notation].  Considering the case $n=8$ gives us $8+1=9$ distinct
likelihood functions, Figure~\ref{ninelikes}.  We see an unanticipated
asymmetry between $r$ and $n-r$.  For $r=4$ one might expect that
$\hat{a}=\frac{1}{2}$ on the grounds that the focal competitor's
performance was median.  However, $\hat{a}\simeq 0.5886$.
Figure~\ref{dotprobs} shows the maximum likelihood estimator $\hat{a}$
for various values of $n,r$, obtained by direct numerical maximization
of equation~\ref{likeforrn3}.

@@

Analysis of expression~\ref{likeforrn3} for fixed $n$ shows that the
support $\mathcal{S}=\log\mathcal{L}_{r,n}(a)$ is asymptotically

\begin{equation}\label{asymptotic}
\mathcal{S}=
%\log\mathcal{L}_{r,n}(a)=
\log(B-1)-B\frac{r+1}{n}
+\mathcal{O}\left(n^{-2}\right)
\end{equation}

(up to an arbitrary additive constant).  Using
$\mathcal{S}^*=\log(B-1)-B\frac{r+1}{n}$ as an approximate
support thus incurs an error of only $\mathcal{O}(n^{-2})$.  Observing
that $\partial^2\mathcal{S}^*/\partial B^2=-(B-1)^{-2}<0$, the evaluate
would be unique; we have
$\hat{B}=\frac{n+r+1}{r+1}+\mathcal{O}\left(n^{-1}\right)$, or
alternatively $\hat{a}=\frac{n}{n+r+1}$.  Compare this with the result
of asking ``what is the probability that the focal competitor beats a
randomly chosen opponent?'' to which the binomial maximum likelihood
estimate is simply $\frac{n-r}{n}$, agreeing to within
$\mathcal{O}(n^{-2})$.  For repeated observations $r_i,n_i$,
$1\leqslant i\leqslant m$ we would have
\begin{equation}
  \hat{a} = 1-
  \frac{\frac{r_1+1}{n_1}+\cdots+\frac{r_m+1}{n_m}}{m}
  = 1-\overline{\left(\frac{r+1}{n}\right)}.
\end{equation}

(that is, one minus the mean ratio of rank to number of [cloned]
competitors); compare the binomial result of $1-\left.\sum
r_i\right/\sum n_i$.

\begin{figure}[t]
  \begin{centering}
\includegraphics[width=4in]{dotprobs}  % produced by very_simplified_likelihood.Rmd
\caption{Maximum likelihood estimator $\hat{a}$ for $n=1(1)10$
  and\label{dotprobs}} $r=0(1)n$
\end{centering}
\end{figure}

Below, I present three examples that showcase likelihood
function~\ref{likeforrn3}, drawn from the fields of athletics,
education, and Formula~1 motor racing.

>>>>

Here a simplification of the Plackett--Luce likelihood functions for
order statistics is proposed.  We consider a field of $n$ entities
that are ordered in some way [the preferred example is competitors in
  a running race, ordered by time of crossing the finishing line].
One entity is of particular interest to us: the ``focal entity'', or
focal competitor.  We assign Plackett--Luce strength $a$ to the focal
competitor and $b$ to all others; we require $a+b=1$.  The associated
inference problem has a wide range of applications and I present some
analysis of datasets drawn from the fields of Olympic athletics,
education, and Formula~1 motor racing.

}
\Keywords{keywords, comma-separated, non-capitalized, \proglang{R}}
\Plainkeywords{keywords, comma-separated, non-capitalized, R} %% without formatting
%% at least one keyword must be supplied

@@

As noted, likelihood function~\ref{plackettluce} becomes unmanageable
with $n$ even moderately large and some form of simplification is
needed.  To that end, consider a Plackett-Luce likelihood function for
a race between a competitor of Bradley-Terry strength $a$ and
$n-1\geqslant 1$ cloned competitors, each of strength $b=1-a$.  Under
these circumstances, an observation is indexed by~$r$, the rank of the
focal competitor, and because we have $n$ competitors $1\leqslant
r\leqslant n$.  The initial field strength is $a+(n-1)b=1+(n-2)b$.  The
Plackett--Luce likelihood function~\citep{luce1959,plackett1975} for
this situation would be

\begin{eqnarray}\label{likeforrn1}
  \mathcal{L}_{r,n}(a) &=&
\frac{b}{1+ (n-2)b}\cdot
\frac{b}{a+(n-3)b}\cdots\frac{a}{a+(n-r-1)b}\nonumber\\
&=& \frac{B-1}{(B+n-1)(B+n-2)\cdots(B+n-r-1)}\nonumber\\ 
&=& \frac{(B-1)(B+n-r-2)!}{(B+n-1)!}\label{likeforrn3}
\end{eqnarray}

\noindent where $B=1/b$.  Note carefully that these expressions are
likelihoods, not probabilities, and do not sum to 1.  Previous related
work \citep{hankin2024_hyper3} presented the case $n=3$ [in current
  notation].  Considering the case $n=8$ gives us $8+1=9$ distinct
likelihood functions, Figure~\ref{ninelikes}.  We see an unanticipated
asymmetry between $r$ and $n-r$.  For $r=5$ one might expect that
$\hat{a}=\frac{1}{2}$ on the grounds that the focal competitor's
performance was median.  However, $\hat{a}\simeq 0.5886$.
Figure~\ref{dotprobs} shows the maximum likelihood estimator $\hat{a}$
for various values of $n,r$, obtained by direct numerical maximization
of equation~\ref{likeforrn3}.

@@

Analysis of expression~\ref{likeforrn3} for large $n$ and $r=pn$ for
fixed $p$ shows that the support
$\mathcal{S}=\log\mathcal{L}_{r,n}(a)$ is asymptotically

\begin{equation}\label{asymptotic}
\mathcal{S}=
%\log\mathcal{L}_{r,n}(a)=
\log(B-1) + B\log(1-r/n)
+\mathcal{O}\left(n^{-1}\right)
\end{equation}

(up to an arbitrary additive constant).  Using
$\mathcal{S}^*=\log(B-1)+B\log(1-r/n)$ as an approximate support thus
incurs an error of only $\mathcal{O}(n^{-1})$.  Observing that
$\partial^2\mathcal{S}^*/\partial B^2=-(B-1)^{-2}<0$, the evaluate
would be unique; we have $\hat{B}=1-\log(1-r/n)^{-1}$, or
alternatively $\hat{a}=(1-\log(1-r/n))^{-1}$.  Compare this with the
result of asking ``what is the probability that the focal competitor
beats a randomly chosen opponent?'' to which the binomial maximum
likelihood estimate is $\frac{n-r}{n}$.  For repeated independent
observations $r_i,n_i$, $1\leqslant i\leqslant N$ we would have

\begin{equation}
  \hat{a} =   \frac{N}{N+\sum\log(1-r_i/n_i)};
\end{equation}

compare the binomial result of $1-\left.\sum r_i\right/\sum n_i$.
Below, I present three examples that showcase likelihood
function~\ref{likeforrn3}, drawn from the fields of athletics,
education, and Formula~1 motor racing.

\begin{figure}[t]
  \begin{centering}
\includegraphics[width=4in]{dotprobs}  % produced by very_simplified_likelihood.Rmd
\caption{Maximum likelihood estimator $\hat{a}$ for $n=2(1)10$
  and\label{dotprobs}} $r=1(1)n$
\end{centering}
\end{figure}
